\section{Quotients by subsets}

Intuitively, quotient space is the space obtained by gluing equivalent points together.
For example, consider the \emph{minimal equivalence relation} on $[0,1]$ such that $0\sim 1$;
that is, $x\sim y$ if and only if one of the following conditions hold $x=y$, or $x=0$ and $y=1$, or $x=1$ and $y=0$.
Then the quotient space $[0,1]/{\sim}$ is homeomorphic to 
\[\mathbb{S}^1=\set{(x,y)\in\RR^2}{x^2+y^2=1}.\]
A homeomorphism is induced by the map $[0,1]\to \mathbb{S}^1$
\[f(t)=\left(\cos(2\cdot\pi\cdot t),\sin(2\cdot\pi\cdot t)\right).\]
The latter statement can be proved directly from the definition of quotient topology, but soon we will prove the following statement that implies this and similar statements effortlessly.
So we suggest to wait with proof of this statement.

\begin{itemize}
\item Suppose $\c{X}$ is a \emph{compact space} and  $\c{Y}$ is \emph{Hausdorff space} and $f\: \c{X}\to\c{Y}$ is a surjective continuous map.
Then $\c{Y}$ is equipped with the quotient topology.
\end{itemize}















In other words, $f\:\c{X}\to \c{Y}$ is an \emph{emebedding} if it is injective and the topology on $\c{X}$ coinsides with pullback topology for $f$.

A function $f\:\c{X}\to \c{Y}$ is called local embedding if for any point $x\in \c{X}$ there is a neighborhood $U\ni x$ such that the restriction $f|_U$ is an embedding.\footnote{The meaning of the word \emph{local} in topology can be described the following way: a space $\c{X}$ is \emph{locally magnificent} if for any point $x\in \c{X}$ and any neighborhood $V\ni x$ there is a smaller neighborhood $W$ (that is, $x\in W\subset V$) which is \emph{magnificent}.}
 
 
 
 
 
 
 
 
 
 
 
 
The following proposition provides a way to construct the weakest topology containing the given collection of sets.

\begin{thm}{Proposition}
Let $\s{Q}$ be arbitrary collection of subsets of a set $\c{X}$.
Consider the set $\s{P}$ of all finite intersections of sets in $\s{Q}$;
that is, $P\in \s{P}$ if there is a finite collection of subsets $Q_1,\dots,Q_n\in \s{Q}$ such that 
\[P=Q_1\cap\dots\cap Q_n.\]
If the collection is empty we assume that $P=\c{X}$.

Denote by $\s{O}$ all possible unions of sets in $\s{P}$;
that is, $O\in \s{O}$ if there is, a (possibly infinite) collection of sets $P_\alpha\in\s{P}$, $\alpha\in\c{I}$ such that 
\[O=\bigcup_{\alpha\in\c{I}}P_\alpha.\]
Then $\s{O}$ is a topology on $\c{X}$.
\end{thm}

\parit{Proof.}
We need to check if the collection $\s{O}$ satisfies the conditions in the definition \ref{def:top-space}.

The conditions \ref{def:top-space:empty} and \ref{def:top-space:u} hold evidently.
It remains to check condition \ref{def:top-space:n};
that is, given two sets $O,O'\in\s{O}$ we need to show that $O_1\cap O_2\in \s{O}$.

Let
\[O=\bigcup_{\alpha\in \c{I}}V_\alpha,\quad O'=\bigcup_{\beta\in \c{J}}V'_\beta,\]
where  $V_\alpha,V'_\beta\in\s{P}$ for any $\alpha\in\c{I}$ and $\beta\in\c{J}$.

Fix $\alpha\in\c{I}$ and $\beta\in\c{J}$.
Since $V_\alpha,\in\s{P}$ it can be presented as an intesection
\[V_\alpha=Q_1\cap \dots \cap Q_n,\]
with $Q_1,\dots,Q_n\in \s{Q}$.
Similarly, 
\[V'_\beta=Q'_1\cap \dots \cap Q'_m,\]
with $Q'_1,\dots,Q'_m\in \s{Q}$.
It follow that 
\begin{align*}
W_{\alpha,\beta}&=V_\alpha\cap V'_\beta=
\\
&=Q_1\cap \dots \cap Q_n\cap Q'_1\cap \dots \cap Q'_m.
\end{align*}
In particular $W_{\alpha,\beta}\in \s{P}$.

Finally note that
\[O\cap O'=\bigcup_{\alpha\in \c{I},\beta\in\c{J}}W_{\alpha,\beta}.\]
Hence $O\cap O'\in \s{O}$.
\qeds
