\chapter{Subsets}

\section{Closed sets}\label{sec:closed-sets}

The definitions of open and closed sets are mirror-symmetric to each other.
There is no particular reason why we define topological space using open sets --- we could use closed sets instead.
In fact, closed sets were considered before open sets --- the former were introdiced by Georg Cantor in 1884 \cite{cantor}, and the latter by René Baire in 1899 \cite{baire}.

Let $\c{X}$ be a topological space. 
A subset $K\subset \c{X}$ is called \index{close subset}\emph{closed} if its complement $\c{X}\setminus K$ is open.

Sometimes it is easier to use closed sets; 
for example, the cofinite topology can be defined by declaring that the whole space and all its finite sets are closed.

From the definition of topological spaces the following properties of closed sets follow.

\begin{thm}{Proposition}\label{prop:closed-properties}
Let $\c{X}$ be a topological space. 
\begin{subthm}{}
 The empty set and $\c{X}$ are closed.
\end{subthm}

\begin{subthm}{}
 The intersection of any collection of closed sets is a closed set.
That is, if $K_\alpha$ is open for any $\alpha$ in the index set $\c{I}$,
then the set
\[Q=\bigcap_{\alpha\in \c{I}} K_\alpha
=
\set{x\in \c{X}}{x\in K_\alpha\quad\text{for any}\quad\alpha\in \c{I}}\]
is closed.
\end{subthm}

\begin{subthm}{def:closed-properties:n}
The union of two closed sets (or any finite collection of closed sets) is closed.  
That is,  if $K_1$ and $K_2$ are closed, then the union $Q=K_1 \cup K_2$ is closed. 
\end{subthm}

\end{thm}

The following proposition is completely analogous to the original definition of continuous maps via open sets (\ref{def:cont-top}). 

\begin{thm}{Proposition}\label{prop:cont-closed}
Let $\c{X}$ and $\c{Y}$ be topological spaces.
A map $f\:\c{X}\z\to\c{Y}$ is continuous if and only if any closed set $Q$ has closed inverse image $f^{-1}(Q)$.
\end{thm}


\parit{Proof.}
In the proof, we will use following set-theoretical identity.
Suppose $A\subset \c{Y}$ and $B=\c{Y}\setminus A$ (or, equvalently, $A=\c{Y}\setminus B$).
Then
$$f^{-1}(B)
=
\c{X}\setminus f^{-1}(A)\eqlbl{eq:inverse-image-complement}$$
for any map $f\:\c{X}\to\c{Y}$.
This identity is tautological,
to prove it observe that both sides can be spelled as 
\[\set{x\in\c{X}}{f(x)\notin A}.\]

\parit{Only-if part.}
Let $B\subset \c{Y}$ be a closed set.
Then $A=\c{Y}\setminus B$ is open.
Since $f$ is continuous, 
$f^{-1}(A)$ is open.
By \ref{eq:inverse-image-complement},
$f^{-1}(B)$ is the complement of $f^{-1}(A)$ in $\c{X}$.
Hence $f^{-1}(B)$ is closed.

\parit{If part.}
Fix an open set $B$, its complement $A=\c{Y}\setminus B$ is closed.
Therefore $f^{-1}(A)$ is closed.
By \ref{eq:inverse-image-complement},
$f^{-1}(B)$ is a complement of $f^{-1}(A)$  in $\c{X}$. 
Hence $f^{-1}(B)$ is open.

The statement follows since $B$ is an arbitrary open set.
\qeds

\begin{thm}{Exercise}\label{ex:closed-continuous}
Let $f\:\c{X}\to \c{Y}$ be a map between topological spaces.
Assume $A,B\subset \c{X}$ are closed sets such that $A\cup B=\c{X}$.
Show that $f$ is continuous if and only if so are the restrictions $f|_A$ and $f|_B$.
\end{thm}


\section{Interior and closure}

Let $A$ be an arbitrary subset in a topological space $\c{X}$.

The union of all open subsets of $A$ is called the \index{interior}\emph{interior} of $A$ and denoted as $\mathring A$.

Note that $\mathring A$ is open.
Indeed, it is defined as a union of open sets and such union has to be open by definition of topology (\ref{def:top-space}).
So we can say that $\mathring A$ is the \textit{maximal} open set in $A$, 
as any open subset of $A$ lies in~$\mathring A$.

The intersection of all closed subsets containing $A$ is called the \index{closure}\emph{closure} of $A$ and denoted as $\bar A$.

The set $\bar A$ is closed.
Indeed, it is defined as an intersection of closed sets and such intersection has to be closed by \ref{prop:closed-properties}.
In other words, $\bar A$ is the \textit{minimal} closed set that contains $A$, 
as any closed subset of $A$ contains $\bar A$.


\begin{thm}{Exercise}\label{ex:closure-interior-complement}
Assume $A$ is a subset of a topological space $\c{X}$;
consider its complement $B=\c{X}\setminus A$.
Show that 
\[\bar B=\c{X}\setminus \mathring A.\]
\end{thm}

The following exercises are a based on the so-called \textit{Kuratowski's problem}.

\begin{thm}{Exercise}\label{ex:closure-closure}
Show that the following holds for any set $A$ of a topological space:
\[\mathring{\mathring A}=\mathring A\subset A\subset \bar A=\bar{\bar A}\]

\end{thm}

\begin{thm}{Exercise}\label{ex:kuratowski-c}
Let $A$ and $B$ be subsets of a topological space.
Suppose that $A\subset B$.
Show that 
\[\bar A\subset \bar B
\quad\text{and}\quad 
\mathring A\subset \mathring B.\]

\end{thm}

\begin{thm}{Exercise}\label{ex:kuratowski-closed}
Show that 
\[\bar{\mathring Q}\subset Q\]
for any closed set $Q$.
\end{thm}

\begin{thm}{Exercise}\label{ex:kuratowski-open}
Show that 
\[\mathring{\bar V}\supset V\]
for any open set $V$.
\end{thm}



\begin{thm}{Advanced exercise}\label{ex:kuratowski}
Give an example of a topological space $\c{X}$ with a subset $A$ such that all the following 7 subsets are distinct:
\[\bar{\mathring{\bar A}},\ \mathring{\bar A},\ \bar A,\  A,\  \mathring A,\ \bar{\mathring A},\ \mathring{\bar{\mathring A}}.\]
Show that given a set $A$ one can get at most 7 by repeatedly applying the set operations of closure and interior.

\end{thm}

\section{Boundary}

Let $A$ be an arbitrary subset in a topological space $\c{X}$.
The \index{boundary}\emph{boundary} of $A$ (briefly $\partial A$) is defined as the complement 
\[\partial A=\bar A\setminus \mathring A.\]

\begin{thm}{Exercise}\label{ex:bry-closed}
Show that the boundary of any set is closed.
\end{thm}

\begin{thm}{Exercise}\label{ex:bry<closure}
Show that the set $A$ is closed if and only if $\partial A\subset A$.
\end{thm}

\begin{thm}{Advanced exercise}\label{ex:3=bry}
Find three disjoint open sets on the real line 
that have the same nonempty boundary.
\end{thm}

\section{Neighborhoods}

Let $x$ be a point in a topological space $\c{X}$.
A \index{neighborhood}\emph{neighborhood} of $x$ is any open set $N$ containing $x$.
In topology, neighborhoods often replace the notion of balls (the latter can be used only in metric spaces).  

\begin{thm}{Exercise}\label{ex:bry-nbhd}
Let $A$ be a set in a topological space $\c{X}$.
Show that $x\in \partial A$ if and only if any neighborhood of $x$ contains points in $A$ and its complement $\c{X}\setminus A$. 
\end{thm}

Let $A$ and $B$ be subsets of a topological space $\c{X}$.
The set $A$ is said to be \index{dense}\emph{dense} in $B$ if $\bar A\supset B$.

\begin{thm}{Exercise}\label{ex:dense-nbhd} 
Show that $A$ is dense in $B$ if and only if any neighborhood of any point in $B$ intersects $A$.
\end{thm}

\begin{thm}{Exercise}\label{ex:nowhere-dense}
Classify topological spaces (up to homeomorphism) containing a unique nowhere
dense subset.
\end{thm}


\section{Limits}

\begin{thm}{Definition}\label{def:limit-top}
Suppose $x_n$ is a sequence of points in a topological space $\c{X}$.
We say that $x_n$ \index{convergence}\emph{converges} to a point $x_\infty\in \c{X}$ (briefly $x_n\to x_\infty$ as $n\to\infty$) if for any neighborhood $N$ of $x_\infty$, we have that 
$x_n\in N$ for all sufficiently large $n$.
\end{thm}

\begin{thm}{Exercise}\label{ex:def:limit-metric=def:limit-top}
Prove that above definition agrees with \ref{def:limit-metric}.
In other words, a sequence of points $x_1,x_2,\dots$ in a metric space converges to a point $x_\infty$ in the sense of definition \ref{def:limit-metric} if and only if it converges in the sense of definition \ref{def:limit-top}.
\end{thm}

\begin{thm}{Exercise}\label{ex:concrete-lim}
Show that in a space with concrete topology any sequence converges to any point.
In particular, a sequence might have different limits.
\end{thm}

\begin{thm}{Exercise}\label{ex:lim-weaker}
Show that a convergent sequence of points in a topological space is also convergent for every weaker topology.
\end{thm}

The following exercise shows that in general, converging sequences do \textit{not} provide an adequate description of topology.
In other words, an analog of \ref{ex:continuous-limit} does not hold.%
(The so-called \index{net}\emph{nets} provide an appropriate generalization of sequences that works well in topological spaces, but we are not going to consider them in the sequel.)

Recall that a set is called \index{countable set}\emph{countable} if it admits a bijection to some subset of the set of natural numbers.
In particular, all finite sets are countable.

\begin{thm}{Advanced exercise}\label{ex:cocountable-top}
Let $\c{X}$ be $\RR$ with the so-called \index{cocountable topology}\emph{cocountable topology};
its closed sets are either countable or the whole $\RR$.

\begin{subthm}{}
 Construct a map $f\:\c{X}\to \c{X}$ that is not continuous.
\end{subthm}

\begin{subthm}{}
 Describe all converging sequences in $\c{X}$.
\end{subthm}

\begin{subthm}{}
 Show that if the sequence $x_n$ converges to $x_\infty$ in $\c{X}$ then for any map $f\:\c{X}\to \c{X}$ the sequence $y_n=f(x_n)$ converges to $y_\infty\z=f(x_\infty)$.
\end{subthm}

\end{thm}
