\chapter{Constructions}

In this chapter we will discuss a few constructions that produce new topological spaces from the given ones.

\section{Induced topology}\label{sec:induced-topology}

Let $A$ be a subset of a topological space $\c{Y}$.
Consider the so-called \emph{induced topology} on $A$ defined the following way:
a subset $V\subset A$ is open in $A$ if and only if $V=A\cap W$ for an open set $W$ in $\c{Y}$.

Let us check that induced topology is indeed a topology;
in other words, it meets all conditions in \ref{def:top-space}.

First of all the empty set
$\emptyset$ is open since $\emptyset=A\cap \emptyset$.
Further, $A=A\cap \c{Y}$;
therefore $A$ is open in the induced topology.

Assume $\set{V_\alpha}{\alpha\in\c{I}}$ is a collection of open sets in $A$;
that is, for each $V_\alpha$ there is a set $W_\alpha$ which is open in $\c{Y}$ and such that $V_\alpha\z=A\cap W_\alpha$.
Note that
\[\bigcup_\alpha V_\alpha=A\cap\left(\bigcup_\alpha W_\alpha\right).\]
Since the union of $\{W_\alpha\}$ is open in $\c{Y}$, the union of $\{V_\alpha\}$ is open in the induced topology on $A$.

Assume $V_1$ and $V_2$ are open in $A$; 
that is, $V_1=A\cap W_1$ and  $V_2\z=A\cap W_2$ for some open sets $W_1$ and $W_2$ in $\c{Y}$.
Note that
\[V_1\cap V_2=A\cap(W_1\cap W_2).\]
Since the intersection $W_1\cap W_2$ is open in $\c{Y}$,
the intersection $V_1\cap V_2$ is open in $A$.


A subset $A$ in a topological space $\c{Y}$ equipped with the induced topology is called a \emph{subspace} of $\c{Y}$.
It is straightforward to check that this notion agrees with the notion introduced in \ref{sec:subspaces-metric};
that is, if $\c{Y}$ is a metric space, then any subset $A\subset\c{Y}$ comes with metric
and the topology defined by this metric coincides with the induced topology on $A$.

A function $f\:\c{X}\to \c{Y}$ is called \emph{embedding} if $f$ defines a homeomorphism from space $\c{X}$ to the subspace $f(\c{X})$ in $\c{Y}$.

\section{Moving topology by a map}

The construction in the following exercise moves topology from target space to the source of a map.

\begin{thm}{Exercise}\label{ex:move-topology:pullback}
Let $f\:\c{X}\to\c{Y}$ be a function between two sets.
Assume $\c{Y}$ is equipped with a topology.
Declare a subset $V\subset\c{X}$  to be open if there is an open subset $W\subset\c{Y}$ 
such that $V=f^{-1}(W)$.
Show that it defines a topology on $\c{X}$.
\end{thm}

The constructed topology on $\c{X}$ is called \emph{pullback} topology.
It generalizes the notion of induced topology above.
Namely, the induced topology on $A\subset \c{Y}$ can be defined as a pullback topology for the inclusion map $\iota\:A\to \c{Y}$.%
\footnote{The inclusion map $\iota\: A\to \c{X}$ is defined by $\iota(a)=a$ for any $a\in A$.}
Indeed, for any $W\subset \c{Y}$ the inverse image $\iota^{-1}(W)$ coincides with the intersection  $V=A\cap W$.

The following exercise describes an analogous construction that moves topology from source to target.
Both exercises can be solved by checking the conditions in \ref{def:top-space} as we did in \ref{sec:induced-topology}.

\begin{thm}{Exercise}\label{ex:move-topology:pushforward}
Let $f\:\c{X}\to\c{Y}$ be a map between two sets.
Assume $\c{X}$ is equipped with a topology.
Declare a subset $W\subset\c{Y}$ to be open if the subset $V=f^{-1}(W)$ is open in $\c{X}$.
Show that it defines a topology on $\c{X}$.

\end{thm}

The constructed topology on $\c{Y}$ is called \emph{pushforward} topology.

\begin{thm}{Exercise}
Let $f\:\c{X}\to\c{Y}$ be a continuous map.

\begin{subthm}{}
Show that the pullback topology on $\c{X}$ is weaker than its own topology.
\end{subthm}

\begin{subthm}{}
Show that the pushforward topology on $\c{Y}$ is stronger than its own topology.
\end{subthm}

\end{thm}

\begin{thm}{Exercise}
Let $g\:\c{X}\to\c{Y}$ be a continuous map.

\begin{subthm}{}
Suppose $\c{X}$ is equipped with the pullback topology.
Show that a map $f\:\c{W}\to\c{X}$ is continuous if and only if the composition $f\circ g\:\c{W}\to\c{Y}$ is continuous.
\end{subthm}

\begin{subthm}{}
Suppose $\c{Y}$ is equipped with the pushforward topology.
Show that a map $h\:\c{Y}\to\c{Z}$ is continuous if and only if the composition $h\circ f\:\c{X}\to\c{Z}$ is continuous.
\end{subthm}

\end{thm}


The pullback topology is used mostly for injective maps; in this case, it is nearly the same as \emph{induced topology}.
Similarly, pushforward topology is mostly used for surjective maps.
This particular case of the construction is called \emph{quotient topology};
it is discussed in the following section.

\begin{thm}{Exercise}\label{ex:open-closed-pushforward}
Let $f\:\c{X}\to \c{Y}$ be a continuous surjective map.
Assume $f$ is closed or open.
Show that $\c{Y}$ is equipped with the quotient topology.
\end{thm}

\section{Quotient topology}

Let $\sim$ be an \emph{equivalence relation} on a topological space $\c{X}$;
that is, for any $x,y,z\in\c{X}$ the following conditions hold:
\begin{itemize}
 \item $x\sim x$;
 \item if $x\sim y$, then $y\sim x$;
 \item if $x\sim y$ and $y\sim z$, then $x\sim z$.
\end{itemize}

Recall that the set 
\[[x]=\set{y\in \c{X}}{y\sim x}\]
is called the \emph{equivalence class} of $x$.
The set of all equivalence classes in $\c{X}$ will be denoted by $\c{X}/{\sim}$.

The following exercise ties equivalence relations with maps.

\begin{thm}{Exercise}
Show that an arbitrary map $f\:\c{X}\to \c{Y}$ defines the following equivalence relation on $\c{X}$:
\[x\sim x'\quad\text{if and only if}\quad f(x)=f(x').\]
Moreover,
\[y=f(x)\quad\text{if and only if}\quad [x]=f^{-1}\{f(x)\}.\]
\end{thm}

Observe that $x\mapsto [x]$ defines a surjective map $\c{X}\to \c{X}/{\sim}$.
The corresponding pushforward topology on $\c{X}/{\sim}$ is called \emph{quotient topology} on $\c{X}/{\sim}$.
By default, $\c{X}/{\sim}$ is equipped with the quotient topology
in this case, it is called \emph{quotient space}.

\section{Quotients by subsets}

Intuitively, quotient space is the space obtained by gluing equivalent points together.
For example, consider the \emph{minimal equivalence relation} on $[0,1]$ such that $0\sim 1$;
that is, $x\sim y$ if and only if one of the following conditions hold $x=y$, or $x=0$ and $y=1$, or $x=1$ and $y=0$.
Then the quotient space $[0,1]/{\sim}$ is homeomorphic to 
\[\mathbb{S}^1=\set{(x,y)\in\RR^2}{x^2+y^2=1}.\]
A homeomorphism is induced by the map $[0,1]\to \mathbb{S}^1$
\[f(t)=\left(\cos(2\cdot\pi\cdot t),\sin(2\cdot\pi\cdot t)\right).\]
The latter statement can be proved directly from the definition of quotient topology, but soon we will prove the following statement that implies this and similar statements effortlessly.
So we suggest to wait with proof of this statement.

\begin{itemize}
\item Suppose $\c{X}$ is a \emph{compact space} and  $\c{Y}$ is \emph{Hausdorff space} and $f\: \c{X}\to\c{Y}$ is a surjective continuous map.
Then $\c{Y}$ is equipped with the quotient topology.
\end{itemize}

Given a subset $A$ in a topological space $\c{X}$, the space $\c{X}/A$ is defined as the quotient space $\c{X}/{\sim}$ for the minimal equivalence relation such that $a\sim b$ for any $a,b\in A$.
For example the quotient space $[0,1]/{\sim}$ discussed above can be also denoted by $[0,1]/\{0,1\}$ --- it is the interval $[0,1]$ with identified two-element subset $\{0,1\}$.

\begin{thm}{Exercise}
Describe the quotient space $[0,1]/(0,1)$, where $[0,1]$ is the real interval with standard topology;
that is, list its points and its open sets.
\end{thm}


\section{Orbit spaces}

\begin{thm}{Definition}
Let $\c{X}$ be a topological space and $G$ be a group.
Suppose that $(g,x)\mapsto g\cdot x$ is a map $G\times\c{X}\to \c{X}$ such that
\begin{subthm}{}
$1\cdot x=x$ for any $x\in \c{X}$, here $1$ denotes the identity element of $G$;
\end{subthm}
\begin{subthm}{}
$g\cdot (h\cdot x)=(g\cdot h)\cdot x$ for any $g,h\in G$ and $x\in \c{X}$;%
\footnote{This condition means that the expression $g\cdot h\cdot x$ makes sense; that is, it does not depend on parentheses.}
\end{subthm}
\begin{subthm}{}
for any $g\in G$, the map $x\mapsto g\cdot x$ is continuous.
\end{subthm}
Then we say that $G$ \emph{acts} on $\c{X}$, or $\c{X}$ is a $G$-space (briefly $G\acts \c{X}$).

In this case, the set
\[G\cdot x:=\set{g\cdot x}{g\in G}\]
is called the \emph{$G$-orbit} of $x$ (or, briefly, \emph{orbit}).
\end{thm}

\begin{thm}{Exercise}
Suppose that a group $G$ acts on a topological space $\c{X}$.
Show that for any $g\in G$, the map $x\mapsto g\cdot x$ defines a homeomorphism $\c{X}\to \c{X}$.
\end{thm}

Suppose that a group $G$ acts on a topological space $\c{X}$.
Set $x\sim y$ if there is $g\in G$ such that $y=g\cdot x$.

Observe that $\sim$ is an equivalence relation on $\c{X}$.
Indeed, 
$x\sim x$ since $x=1\cdot x$.
Further, if $y=g\cdot x$, then 
\[x=1\cdot x=g^{-1}\cdot g\cdot  x=g^{-1}\cdot y;\] since $g^{-1}\in G$ we get that $x\sim y$
$\Longrightarrow$
$y\sim x$.
Finally, suppose $x\sim y$ and $y\sim z$;
that is, $y=g\cdot x$ and $z=h\cdot y$ for some $g,h\in G$.
Then $z=h\cdot g\cdot x$;
therefore $x\sim z$.

For the described equivalence relation, the quotient space $\c{X}/{\sim}$ can be also denoted by $\c{X}/G$;
it is called quotient of $\c{X}$ by the action of~$G$.

Note that $[x]=G\cdot x$; that is, the orbit of $x$ coincides with its equivalence class.
By that reason $\c{X}/G$ is also called \emph{orbit space}.

\begin{thm}{Exercise}
Suppose a group $G$ acts on a topological space $\c{X}$ and $f\:\c{X}\to \c{X}/G$ is the quotient map.

\begin{subthm}{}
Show that $f$ is open.
\end{subthm}

\begin{subthm}{}
Assume $G$ is finite.
Show that $f$ is closed.
\end{subthm}

\end{thm}

