\chapter{Topological spaces}

%Topology studies \emph{topological spaces} --- sets with certain structure that is also named \emph{topology}.
%This structure makes possible to talk about \emph{continuous} maps.

In the previous chapter we defined open sets in metric spaces 
and showed that continuity could be defined using only the notion of open sets.
Now we will state the most important properties of these open sets as axioms.
It will give us a definition of \emph{topological space} as a set with a distinguished class of subsets called \emph{open sets}.

The \emph{topological properties} are loosely defined as properties which survive under arbitrary continuous deformation.
They were studied since 19th century.
The first definition of topological spaces was given by Felix Hausdorff in 1914.
In 1922, the definition was generalized slightly by Kazimierz Kuratowski; his definition is given below.

\section{Definitions}

We are about to define \emph{abstract open sets} without referring to metric spaces.
The exercises \ref{ex:open-union} and \ref{ex:open-intersection} motivate this definition.

\begin{thm}{Definition}\label{def:top-space}
Suppose $\c{X}$ is a set 
with a distinguished class of subsets, called \an{open sets} such that

\begin{subthm}{def:top-space:empty}
The empty set $\emptyset$ and the whole $\c{X}$ are open.
\end{subthm}

\begin{subthm}{def:top-space:u}
The union of any collection of open sets is an open set.
That is, if $V_\alpha$ is open for any $\alpha$ the index set $\c{I}$, 
then the set
\[W=\bigcup_{\alpha\in \c{I}} V_\alpha
=
\set{x\in \c{X}}{x\in V_\alpha\quad\text{for some}\quad\alpha\in \c{I}}\]
is open.
\end{subthm}

\begin{subthm}{def:top-space:n}
The intersection of two open sets is an open set.  
That is, if $V_1$ and $V_2$ are open, then the intersection $W=V_1 \cap V_2$ is open. 
\end{subthm}

In this case, $\c{X}$ is called \an{topological space}.

The collection of all open sets in  $\c{X}$ is called a \an{topology} on $\c{X}$ and denoted as $\s{O}_{\c{X}}$;
so instead of saying \emph{$V$ is an open set in the topological space $\c{X}$}, we might write $V\in \s{O}_{\c{X}}$.
\end{thm}

From (\ref{def:top-space:n}) it follows that the intersection of a finite collection of open sets is open.
That is, if $V_1$, $V_2$, $\dots, V_n$ are open, then the intersection 
\[W=V_1 \cap V_2\cap\dots\cap V_n\] is open.
This can be proved by applying induction on $n$ since 
\[V_1 \cap \dots\cap V_{n-1}\cap V_n=(V_1 \cap \dots V_{n-1})\cap V_n.\]

\section{Examples}

The so-called \emph{connected two-point space} is a simple but nontrivial example of topological space.
This space consists of two points 
\[\c{X}=\{a,b\}\]
and it has three open sets: 
\[\emptyset,\quad \{a\}\quad\text{and}\quad\{a,b\}.\]
It is instructive to check that this is indeed a topology.

Further, for any set $\c{X}$, we can always define the following topologies:
\begin{itemize} 
\item  The \emph{discrete topology} --- the topology consisting of all subsets of a set $\c{X}$.
\item  The \emph{concrete topology}  --- the topology consisting of just whole space $\c{X}$ and the empty set, $\emptyset$.
\item  The \emph{cofinite topology} --- the topology consisting of the empty set, $\emptyset$ and the complements to finite sets.
\end{itemize}

\begin{thm}{Exercise}\label{ex:Rge}
Show that $\emptyset$, $\RR$ and the intervals $[a,\infty)$, $(a,\infty)$ for all $a\in\RR$ define a topology on the real line $\RR$.
(The obtained space will be denoted by $\RR_\ge$.)
\end{thm}

\section{Comparison of topologies}

Let $\s{W}$ and $\s{S}$ be two topologies on one set.
Suppose $\s{W}\subset\s{S}$; that is, any open set in $\s{W}$-topology is open in $\s{S}$-topology.
In this case we say that $\s{W}$ is \emph{weaker} than $\s{S}$, or, equivalently, $\s{S}$ is \emph{stronger} than~$\s{W}$.

\begin{thm}{Exercise}
Let $\s{W}$ and $\s{S}$ be two topologies on one set.
Suppose that for any point $x$ and any $W\in\s{W}$ such that $W\ni x$, there is $S\in \s{S}$ such that 
$W\supset S\ni x$.
Show that $\s{W}$ is weaker than $\s{S}$.
\end{thm}



\section{Continuous maps}

Our next challenge is to reformulate definitions from the previous chapter using only open sets.
Continuous maps are first in the line.
The following definition is motivated by Proposition~\ref{prop:cont-open}.

\begin{thm}{Definition}\label{def:cont-top}
A function between topological spaces 
$f\:\c{X}\to\c{Y}$ is called \emph{continuous} if for any open set $W$ in $\c{Y}$, its inverse image $f^{-1}(W)$ is open in $\c{X}$.
That is, if $W$ is an open subset in $\c{Y}$, then the set
\[V=f^{-1}(W)=\set{x\in X}{f(x)\in W}\]
is an open subset $\c{X}$
\end{thm}

\begin{thm}{Exercise}
Let $\RR$ be the real line with the standard topology
and $\c{X}$ be the connected two-point space described above.

\begin{subthm}{}
Construct a nonconstant continuous function $\RR\to \c{X}$.
\end{subthm}

\begin{subthm}{}
Show that any continuous function $\c{X}\to \RR$ is constant.
\end{subthm}

\end{thm}

\begin{thm}{Exercise} 
Show that a function $f\:\RR\to\RR$ is nondecreasing if and only if it defines a continuous map $\RR_\ge\to \RR_\ge$.
(The space $\RR_\ge$ is defined in \ref{ex:Rge}.)
\end{thm}

\section{Limits}

\begin{thm}{Definition}\label{def:limit-top}
Suppose $x_n$ is a sequence of points in a topological space $\c{X}$.
We say that $x_n$ \emph{converges} to a point $x_\infty\in \c{X}$ (briefly $x_n\to x_\infty$ as $n\to\infty$) if for any open set $V\ni x_\infty$ there is $N$ such that 
$x_n\in U$ for any $n\ge N$.
\end{thm}

\begin{thm}{Exercise}
Prove that above definition agrees with \ref{def:limit-metric}.
In other words, if $x_1,x_2,\dots$, and $x_\infty$ are points in a metric space, then $x_n$ converges to $x_\infty$ in the sense of definition \ref{def:limit-metric} if and only if $x_n$ converges to $x_\infty$ in the sense of definition \ref{def:limit-top}.
\end{thm}

\begin{thm}{Exercise}
Show that in a space with concrete topology any sequence converges to any point.
In particular, the limit point of a sequence is not uniquely defined.
\end{thm}

\begin{thm}{Exercise}
Show that a convergent sequence of points in a topological space is also convergent for every weaker topology.
\end{thm}

The following exercise shows that in general, converging sequences do \emph{not} provide an adequate description of topology.
In other words, an analog of \ref{ex:continuous-limit} does not hold.%
\footnote{The so-called \emph{nets} provide an appropriate analog of sequences that works well in topological spaces, but we are not going to consider them here.}

\begin{thm}{Advanced exercise}
Let $\c{X}$ be $\RR$ with the so-called \emph{cocountable topology};
its closed sets are either countable or the whole $\RR$.

\begin{subthm}{}
 Construct a map $f\:\c{X}\to \c{X}$ that is not continuous.
\end{subthm}

\begin{subthm}{}
 Describe all converging sequences in $\c{X}$.
\end{subthm}

\begin{subthm}{}
 Show that if the sequence $x_n$ converges to $x_\infty$ in $\c{X}$ then for any map $f\:\c{X}\to \c{X}$ the sequence $y_n=f(x_n)$ converges to $y_\infty\z=f(x_\infty)$.
\end{subthm}

\end{thm}

\section{Metrizable spaces}

According to exercises \ref{ex:open-union} and \ref{ex:open-intersection} any metric space is a topological space if one defines open sets as in the definition \ref{def:open}.
As it follows from Exercise \ref{ex:d1+d2+dinfty-open},
different metrics on one set might define the same topology.

A topological space is called \emph{metrizable} if its topology can be defined by a metric.
Let us give few examples of nonmetrizable spaces.

\begin{thm}{Exercise}\label{ex:finite+metrizable}
Show that finite topological space is metrizable if and only if it is discrete.
In particular, connected two-point space is not metrizable.
\end{thm}

\begin{thm}{Exercise}
Assume an infinite set $\c{X}$ equipped with the cofinite topology.
Show that $\c{X}$ is not metrizable.
\end{thm}

\begin{thm}{Exercise}
Show that $\RR_\ge$ is not metrizable.
(The space $\RR_\ge$ is defined in \ref{ex:Rge}.)
\end{thm}
